---
title: "**Homework 1**"
author: "Santiago Ruiz, Nikita Karetnikov, "
output:
  pdf_document:
    number_sections: true
date: "`r Sys.Date()`"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 1

In this exercise, we are asked to calculate the limits for a system of number
representations with the following parameters:


- \( b = 10 \) base
- \( m = 3 \) mantissa length
- \( e_{\min} = -3 \)
- \( e_{\max} = 4 \)


To calculate the required numbers we would need the following formula:

$$
x = (-1)^{w_0} b^e \sum_{i=1}^{m} u_i b^{-i}
$$
Let us implement it in R:

```{r}

calculate <- function(w0, b, e, u,m){
  mantissa <- 0   
  
  for (i in 1:m) {
    mantissa <- mantissa + u[i] * b^(-i)
  }
  
  x <- (-1)^w0 * b^e * mantissa
  
  return(x)
}

```

Good! Now we will be able to determine the required numbers.



## Determine the largest floating point number

We start with  the largest floating point number.
We assume that the largest number should have:

- the largest possible exponent
- mantissa with the largest digits in this system (which should be equal to b-1)


```{r}

b <- 10  
m <- 3

e_min <- -3  
e_max <- 4


x_max <- calculate(w0=0, b=b, e=e_max, u= rep(b-1, m),m=m)
print(x_max)

```



## Determine the smallest positive floating point number

Now we should determine the smallest positive floating point number.
We assume that the largest number should have:

- the smallest possible exponent
- mantissa with the smallest digits in this system (which should be equal to 0)

But since mantissa should start with non-zero we set its digits to [1,0,0] :

```{r}

x_min <- calculate(w0 = 0, b = b, e = e_min, u = c(1,0,0), m = m)

print(x_min)

```

## Determine the largest floating point number smaller than one.

Now we should find the number that approaches value of 1, but that is smaller 
then one.

We set:
- exponent to 0
- mantissa to have largest digits in this system (which should be equal to b-1)


```{r}
x_max_2 <- calculate(w0 = 0, b = b, e = 0, u= rep(b-1, m), m = m)
print(x_max_2)
```

## Determine the smallest floating point number greater than one

Now similarly we should find the number that is closest to one, but is bigger
than one. 

We set:
- exponent to 0
- mantissa digits to be equal to [1,0,1] 

```{r}
x_min_2 <- calculate(w0 = 0, b = b, e = 1, u = c(1,0,1), m = m)
print(x_min_2)

```


# Exercise 2

In this exercise we need to understand the limits or R number representation
system by running experiments with convergence of the series.

## For which n does the loop stop? (practical part)

We first implement convergence in R. We run the loop until the difference
between \( S_{n+1} \) and \( S_n \) becomes smaller than the smallest number
that R can distinguish.
.

```{r}

S_n_prev <- -1  
S_n <- 0
n <- 0  

while (abs(S_n - S_n_prev) > .Machine$double.eps){ 
  n <-  n + 1
  S_n_prev  <- S_n
  S_n <- S_n_prev + 2^(-2 * n)
}

cat("Loop stops at n =",n)

```

## For which n does the loop stop? (theoretical part)

Now we need to theoretically approximate the value of \( n \) at which the loop stops. 
We observe that the convergence of our series resembles the formula used to represent numbers in R.


$$
\sum_{i=1}^{\infty} 2^{-2i} 
\quad\quad\quad\quad
x = (-1)^{w_0} b^{e} \sum_{i=1}^{m} u_i b^{-i}
$$
Let us remind ourselves what is mantissa length and base in  R is:
```{r}


m <-  .Machine$double.digits
cat("Mantissa length (m):", m, "\n")

b <-  .Machine$double.base
cat("Base (b):", b, "\n")

```

With each iteration, the summation approaches its limit more closely, but the increment decreases rapidly (each term is \(2^{-2i}\)). Due to floating-point representation limits in R (with a mantissa length \( m = 53 \) bits for double precision), the smallest distinguishable difference around 1 is about \(2^{-53}\).

The summation stops updating numerically when the next term \(2^{-2n}\) becomes smaller than this limit:

$$
2^{-2n} \approx 2^{-m} \quad \Rightarrow \quad n = \frac{m}{2} = \frac{53}{2} = 26.5
$$
26.5 is quite close to the practical stopping point we found earlier.






# Exercise 3

In this exercise we are asked to approximate exp(x) using Taylor series.


$$
\exp(x) = \sum_{i=0}^{\infty} \frac{x^i}{i!}
$$

## (i)

We implement an algorithm in R that approximates the exponential function $\exp(x)$ using the Taylor series expansion. The algorithm iteratively adds terms from the Taylor series until the absolute value of the next summand is smaller than \texttt{.Machine\$double.eps\textasciicircum(1/2)} times the absolute value of the current approximation.

We run it to identify how many summands we need to approximate $\exp(x=10)$

```{r}




calculate_exp <- function(X){
  term <- 1              
  sum <- term            
  n <- 0                 
  epsilon <- .Machine$double.eps^(1/2)  
    
    
  while (abs(term) > epsilon * abs(sum)) {
    
    n <- n + 1
    term <- (X^n) / factorial(n)  
    sum <- sum + term
  

  } 
  return(sum)
  
}

X <- 40



cat("Our function:", calculate_exp(X), "\n")

cat("R actual function", exp(X), "\n")


```

Looks that our implemented function closely matches R's built-in calculation of the exponential function.
the exponent.


## (ii)


Next, we evaluate the approximation for various values of $X$, both positive
and negative. We calculate the MAPE for values ranging from $-20$ to $200$.






```{r}

X_values <- seq(-20, 200, by=1)
errors <- numeric(length(X_values)) 

for (i in seq_along(X_values)) {
  X <- X_values[i]
  
  
  mape <- mean(abs((exp(X) - calculate_exp(X)) / exp(X))) * 100

  
  
  errors[i] <- mape  
}


plot(X_values, errors, type = "b", col = "blue",
     xlab = "X Value", ylab = "Error",
     main = "Error of Taylor Approximation of exp(X)",
     ylim = c(0, 10))  # Adjust these limits as needed for your data
grid()

```

We notice that for lower values, MAPE increases starting approximately from $-10$.  
For values of $X$ higher than $100$, MAPE is absent as the approximation
of the exponent is probably equal to $\infty$.



Let us check:

```{r}
X <- 101

cat("Our function:", calculate_exp(X), "\n")

cat("R actual function", exp(X), "\n")


```

```{r}
X <- -20

cat("Our function:", calculate_exp(X), "\n")

cat("R actual function", exp(X), "\n")


```



Indeed, either our approximations are too far from the actual values of the
function, or our approximations are equal to infinity.





## (iii)



Let us suggest a modification how to fix the numerical instability. We use the 
following property:


$$
e^{x} = \left(e^{x/n}\right)^n
$$

We implement it in our updated function as a wrapper:




```{r}
calculate_exp_2 <- function(x, n = 10) {
    x_n <- x / n
    result <- calculate_exp(x_n)^n
    return(result)
}


```

Let's first check it for single values:

```{r}
X <- -20

cat("Our function:", calculate_exp_2(X), "\n")

cat("R actual function", exp(X), "\n")


```

```{r}
X <- 101

cat("Our function:", calculate_exp_2(X), "\n")

cat("R actual function", exp(X), "\n")


```

And now for our range:

```{r}

X_values <- seq(-20, 200, by=1)
errors <- numeric(length(X_values)) 

for (i in seq_along(X_values)) {
  X <- X_values[i]
  
  
 mape <- mean(abs((exp(X) - calculate_exp_2(X)) / exp(X))) * 100

  
  
  errors[i] <- mape  
}

# plotting

plot(X_values, errors, type = "b", col = "blue",
     xlab = "X Value", ylab = "Error",
     main = "Error of Taylor Approximation of exp(X)",
     ylim = c(0, 10))  # Adjust these limits as needed for your data
grid()



```


All looks good!


Exercise 4

```{r, include=TRUE, echo=TRUE}

function_x <- function(x) {
  return((sqrt(1 + x) - 1) / x)
}

function_x_rationalisation <- function(x) {
  return(1 / (sqrt(1 + x) + 1))
}

# Determine the values of R for x =1, 10^-4, 10^-32

function_x(1)
function_x(10^(-4))
function_x(10^(-32))

# The Taylor series expansion at 0 is

taylor_expansion_at_zero <- function(y) {
  return(1 + y / 2 - y^2 / 8 + y^3 / 16 - 5 * y^4 / 128 + 7 * y^5 / 256)
}

approximation_result_function_x_at_zero <- function(x) {
  return((taylor_expansion_at_zero(x) - 1) / x)
}

approximation_result_function_x_at_zero(1)
approximation_result_function_x_at_zero(10^(-4))
approximation_result_function_x_at_zero(10^(-32))


# Create a loop that stores the results of a multiplication in a vector
multiplication_results_negative <- numeric(100)
multiplication_results_positive <- numeric(100)

for (i in seq_len(200)) {
  multiplication_results_negative[i] <- .Machine$double.eps*i + 0
  multiplication_results_positive[i] <- -.Machine$double.eps*i + 0
}

combined_results <- c(multiplication_results_negative, multiplication_results_positive)
sorted_results <- sort(combined_results)

print(sorted_results)

function_x_values <- sapply(sorted_results, function_x)
approximation_values <- sapply(sorted_results, approximation_result_function_x_at_zero)
function_x_rationalisation_values <- sapply(sorted_results, function_x_rationalisation)

plot(sorted_results, function_x_values, type = "l", col = "blue", ylim = range(c(function_x_values, approximation_values)), ylab = "Values", xlab = "x", main = "Comparison of function_x and its Taylor Approximation")
lines(sorted_results, approximation_values, col = "red")
lines(sorted_results, function_x_rationalisation_values, col = "green")
legend("topright", legend = c("function_x", "Taylor Expansion", "Function_x_rationalisation"), col = c("blue", "red","green"), lty = 1)


# With greater values of around o
increment_values <- seq(-0.0005, 0.0005, length.out = 1000)

# Calculate function values for the new vector
function_x_increment_values <- sapply(increment_values, function_x)
approximation_increment_values <- sapply(increment_values, approximation_result_function_x_at_zero)
function_x_rationalisation_increment_values <- sapply(increment_values, function_x_rationalisation)

# Plot the new values
plot(increment_values, function_x_increment_values, type = "l", col = "blue", ylim = range(c(function_x_increment_values, approximation_increment_values)), ylab = "Values", xlab = "x", main = "Comparison of function_x and its Taylor Approximation")
lines(increment_values, approximation_increment_values, col = "red")
lines(increment_values, function_x_rationalisation_increment_values, col = "green")
legend("topright", legend = c("function_x", "Taylor Expansion", "Function_x_rationalisation"), col = c("blue", "red", "green"), lty = 1)

print(function_x_increment_values[1])
print(approximation_increment_values[1])
print(function_x_rationalisation_increment_values[1])


```
As shown in the first graph, the divergencews between the values of the taylor
approximation, and the function in its two forms differ when the
values of zero are very small of the order of .Machine$double.eps that
represents the minimal difference stored in doubles so that 1 + .Machine$double.eps is
recognized as different than 1.  According to what we see in class we can think
on two possible answers similar to computing the mean either through the sum of squares
of deviations or as the differences between the mean of the squared values and the 
squared value of the mean, it may be that for values that R uses
different process to solve (sqrt) and polynomial fractions that yield
different results for very small values of x. It can be also that the taylor
approximation is not accurate enough for very small values of x.


* Exercise 5

```{r, include=TRUE, echo=TRUE}

data("KNex", package = "Matrix")
X <- as.matrix(KNex$mm)
Y <- KNex$y

# Not accurate type of execution.
start_time <- Sys.time()
fit <- lm.fit(X, Y)
end_time <- Sys.time()
execution_time_lmfit <- end_time - start_time
print(paste("Execution time:", execution_time))

summary(fit)
print(fit$coefficients)

# Formula of Least Square estimator B = (X^T X)^-1 X^T Y
# Let us try to find this (X^T X)^-1 part with 

###################################################
### Cholesky decomposition
###################################################
# We find the product of the matrix X^TX (This matrix is symetric)
# Crossprod tells him just to do half of the work, by computing only the upper triangular part of the matrix.



R <- chol(crossprod(X))

# Using Solve to find the inverse of both matrix and to multiply the right upper triangular matrix

start_time <- Sys.time()
chol_betas <- backsolve(R, forwardsolve(R, crossprod(X, Y), upper.tri = TRUE, transpose = TRUE))
end_time <- Sys.time()
execution_time_chol <- end_time - start_time
print(paste("Execution time:", execution_time_chol))

# It requires more time than the lm.fit function.


###################################################
### QR decomposition
###################################################
QR <- qr(X)
qr_betas <- backsolve(qr.R(QR), crossprod(qr.Q(QR), Y))
solve(QR, y)

# The betas are identical, but the execution time is different.
fit$coefficients[2]
chol_betas[2]
qr_betas[2]

## 2.nd Part

object.size(X)
X_not_sparsed <- KNex$mm
object.size(X_not_sparsed)
# The class is a dgCMatrix, which is a sparse matrix. Meaning that it 
# saves some space by not saving the zeros.
class(KNex$mm)
class(X)

?uniroot()

```

